The Key to a Performant Data Parallel Compiler

### 2016-11-17 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
More adjustment to the Republication policy stuff.

In the following sections we extend, expound, and clarify on existing work \cite{hsu2016key} to describe the core methods and idioms surrounding the AST encoding, construction of the Path Matrix, and the use of these structures in handling nested, recursive AST relationships, which enable us to perform arbitrary computation over arbitrary sub-trees of an AST selected by parent-child relationships in the manner often seen in compiler transformations. These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}. The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives. To illustrate the performance of these techniques, we benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers. This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.

@@

\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes.

>>>>

These ideas have been further implemented and tested through the implementation of a complete data-parallel by construction compiler, called Co-dfns, that compiles a lexically scoped, functionally oriented dialect of APL with nested functions \cite{hsu2014co,hsu2015accelerating}. The compiler targets both the GPU and the CPU, and its core is implemented in this pure, restricted language that uses only function composition over data-parallel primitives. We extend and improve on prior work \cite{hsu2016key} by simplifying the core expressions and examining specific case studies in greater detail than previously done. We extend the methodology with performance-centered optimizations of the Path Matrix and benchmark a transformation heavy expression lifting algorithm that is found almost universally in compilers. This benchmark shows performance speedups of up to a factor of 58 when compared against pure, functionally oriented algorithms.

@@

\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and report on the results, including the overall architecture of the compiler, the uses of these techniques within the compiler, and the demonstration of these techniques applied to two specific compiler passes in full detail.
